How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
minor changes

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that these visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To that end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of these projects in three different dimensions. Firstly, we look at the types of blocks used and the size of the projects. We then investigate complexity, used abstractions and programming concepts. Finally we detect \emph{code smells} like long scripts and duplicated code blocks. Our results show that 1) most Scratch program are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs suffer do commonly from code smells including long methods and unmatched broadcast signals.
\end{abstract}

@@

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively. Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common. Specifically, the research questions that we are trying to answer are:

@@

Our study is based on data from the Scratch project repository. By scraping the list of recent projects \footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them. To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features and identified smells (Section \ref{sec:results})}
	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

Scratch is a block-based programming language aimed at children, developed by MIT. Scratch can be used to create games and interactive animations, and is available both as a stand-alone application and as a web application. Figure \ref{fig:ui} shows the Scratch user interface in the Chrome browser. The main concepts in the Scratch programming environment are:

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

@@

\subsection{Data analysis}
\label{dataAnalysis}
All scraped project data, including the list of used blocks and parameters, was imported to a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. This data was then imported to spreadsheets for the calculation of statistics and the generation of graphical representations. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the projects.

@@

To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artefacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Histogram of the number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Histogram of the number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Histogram of the number of blocks in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

@@

To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of linearly independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

\subsection{Programming Abstractions and Concepts}
\label{RQ2}
The first method for abstraction that we investigate is the use of procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
	\caption{Histogram of the number of procedures for the 17,979 projects that include at least one}
	\label{fig:proceduresperproject}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/5arguments}
	\caption{Histogram of the number of arguments for the procedures in the dataset}
	\label{fig:procedurearguments}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/7procedurecalls}
	\caption{Histogram of the number of calls of each procedure in the dataset}
	\label{fig:procedurecalls}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
	\caption{Histogram of the number of variables and lists used in the projects}
	\label{fig:variables}
\end{figure}

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ2: One third of the projects use variables, sometimes without initializing them. 39.81\% of the projects contain conditional statements and 77.18\% contain loops, but conditional loops are rarely used. More than half of the projects are interactive. 29.57\% of the projects use broadcast and receive blocks. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
	\end{minipage}
}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
	\caption{Histogram of the number of cloned scripts}
	\label{fig:clonesprojects}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
	\caption{Histogram of the number of copies of the identified clones}
	\label{fig:clonescopies}
\end{figure}

@@

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronizations occurs in 18,669 (7.99\% of the total) projects.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites. 28.16\% of the projects contain code that is never invoked, and thus exhibit the dead code smell. In some cases these scripts are large. The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
	\end{minipage}}

@@

In a sense, the dead scripts are harmless, as they are not executed, however, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast with other visual educational languages, Scratch does on indicate scripts are dead. LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.

Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily. We envision that would be like the `backpack' meant to move scripts across sprites. In order to help novice programmers to keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.

\subsubsection{Exact clones between sprites}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of scripts and procedures}
Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects. This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started quicker.

\subsection{Threads to validity}
A thread to validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects. It could be the case that programming habits of Scratch users are changing over time. However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch programs.\footnote{\url{https://scratch.mit.edu/statistics/}}.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large scale study on \nAnalyzedPrograms~projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.

>>>>

\maketitle
\begin{abstract}
Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education. There is substantial research showing that these visual programming environments are suitable for teaching programming concepts. But, what do people do when they use Scratch? In this paper we explore the characteristics of Scratch programs. To this end we have scraped the Scratch public repository and retrieved 250.000 projects. We present an analysis of these projects in three different dimensions. Initially, we look at the types of blocks used and the size of the projects. We then investigate complexity, used abstractions and programming concepts. Finally we detect \emph{code smells} like large scripts, dead code and duplicated code blocks. Our results show that 1) most Scratch programs are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs do suffer from code smells including large scripts and unmatched broadcast signals.
\end{abstract}

@@

Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}. Research works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}. Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}. A recent controlled experiment found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts. Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common. Specifically, the research questions that we are trying to answer are:

@@

Our study is based on data from the Scratch project repository. By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed analysis on them. To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features, and identified smells (Section \ref{sec:results})}
	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

Scratch is a block-based programming language aimed at children, developed by MIT. Scratch can be used to create games and interactive animations, and is available both as a stand-alone and as a web application. Figure \ref{fig:ui} shows the Scratch user interface. The main concepts in the Scratch programming environment are:

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

@@

\subsection{Data analysis}
\label{dataAnalysis}
All scraped project data, including the list of used blocks and parameters, was imported in a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the scripts.

@@

To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
	\caption{Number of sprites in the analyzed projects}
	\label{fig:size-sprites}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
	\caption{Number of scripts in the analyzed projects}
	\label{fig:size-scripts}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
	\caption{Number of blocks in the analyzed projects}
	\label{fig:size-locs}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
	\caption{McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
	\label{fig:cyclomatic}
\end{figure}

@@

To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

\subsection{Programming Abstractions and Concepts}
\label{RQ2}
The first method for abstraction that we investigate are procedures. In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
	\caption{Number of procedures for the 17,979 projects that include at least one}
	\label{fig:proceduresperproject}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/5arguments}
	\caption{Number of arguments for the procedures in the dataset}
	\label{fig:procedurearguments}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/7procedurecalls}
	\caption{Number of calls of each procedure in the dataset}
	\label{fig:procedurecalls}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
	\caption{Number of variables and lists used in the projects}
	\label{fig:variables}
\end{figure}

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ2: One third of the projects use variables, sometimes without initializing them. 39.81\% of the projects contain conditional statements and 77.18\% contain loops, but conditional loops are rarely used. More than half of the projects are interactive. 29.57\% of the projects use broadcast and receive blocks. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist in 1,052 (0.45\% of the total) projects.
	\end{minipage}
}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
	\caption{Cloned scripts in the dataset projects}
	\label{fig:clonesprojects}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
	\caption{Number of copies of the identified clones}
	\label{fig:clonescopies}
\end{figure}

@@

The second smell that we examine is the dead code smell. We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts. Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects. This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects. For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received. This lack of syncronization occurs in 18,669 (7.99\% of the total) projects.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites. 28.16\% of the projects contain code that is never invoked, and thus exhibit the dead code smell. In some cases these scripts are large. The large script smell is found in 29.77\% of the projects and the large sprite one in 13.68\%.
	\end{minipage}}

@@

In a sense, the dead scripts are harmless, as they are not executed. However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast to other visual educational languages, Scratch does on indicate scripts that are dead. LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.

Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily. We envision that would be like the `backpack' meant to move scripts across sprites. In order to help novice programmers keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.

\subsubsection{Exact clones between sprites}
With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common. In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them. So in many cases there is no way to share the functionality other than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

\subsubsection{Sharing of scripts and procedures}
Investigating the use of clones between projects, we observe that there are 1,700 scripts that are used in multiple projects, sometimes as often as in 1,600 different projects. This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library. An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'. This might empower new Scratch users to get started faster.

\subsection{Threats to validity}
A threat to the validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects. It could be the case that the programming habits of Scratch users are changing over time. However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch projects.\footnote{\url{https://scratch.mit.edu/statistics/}}.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large-scale study on 247,798 projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially from dead code and code duplication.
